---
title: "Additional LoCoh & BB code"
author: "Dana Seidel"
date: "1/8/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Here are some additional notes and code for about Local Covenx Hull and Brownian Bridge methods
for home range analysis in the `move` and `adehabitatHR` packages. 

```{r warnings=FALSE, messages=FALSE}
library(move)
library(adehabitatHR)
library(sp)

data(bear) # load the data
xy <- SpatialPoints(na.omit(ld(bear)[,1:2])) #transform the ltraj to dataframe, remove nas, make spatial points. 

```

# Brownian Bridge Methods
## Kernel BB
```{r}
#sig1	
#first smoothing parameter for the brownian bridge method (related to the speed of the animals; it can be estimated by the function liker).
#sig2	
#second smoothing parameter for the brownian bridge method (related to the imprecision of the relocations, supposed known).

# estimate sig1
liker(bear, sig2 = 30, rangesig1 = c(1, 10))

kud <- kernelbb(bear, sig1 = 3.3153, sig2 = 30)
image(kud) + title("Bear bb_UD")

## Kernel home range
jj <- kernel.area(kud)                  ## home range size
jj
ver95 <- getverticeshr(kud) ## home-range contours
ver80  <- getverticeshr(kud, percent = 80)
plot(ver95)  + plot(ver80, add=TRUE, col="green")  +  points(xy)   ## Plots contours
```

## Dynamic Brownian Bridge UD
```{r}
library(move)
bear_move <- move(x=bear[[1]]$x, y=bear[[1]]$y, time=bear[[1]]$date, proj="+init=epsg:3857")

BB.bear <- brownian.bridge.dyn(bear_move, 
                                dimSize=150, 
                                location.error=30,  #can also provide vector with changing positional errors
                                margin=21,
                                ext=.5,
                                window.size=99, 
                                time.step=2) 
udbear <- getVolumeUD(BB.bear)

plot(bear_move, col="#00000060", pch=16, cex=0.5,
     bty="n", xaxt="n", yaxt="n", xlab=NA, ylab=NA) +
lines(bear_move, col="#00000030") +
# add isopleths
contour(udbear, levels=c(0.5, 0.95), add=TRUE, lwd=c(2, 1), lty=c(2,1), col = c("orange", "blue"))+
  title("Dynamic brownian bridge")
```
Much much much tighter contours than kUD and MCP methods!

## Local-Convex-Hull Methods 

From the adehabitatLT vignette:

> The LoCoH (Local Convex Hulls) family of methods has been proposed by Getz
et al. (2007) for locating Utilization Distributions from relocation data. Three
possible algorithms can be used: Fixed k LoCoH, Fixed r LoCoH, and Adaptive
LoCoH. The three algorithms are implemented in adehabitatHR. All the
algorithms work by constructing a small convex hull for each relocation and its
neighbours, and then incrementally merging the hulls together from smallest to
largest into isopleths. The 10% isopleth contains 10% of the points and represents
a higher utilization than the 100% isopleth that contains all the points (as
for the single linkage method). The three methods are as follows: 
(1) Fixed k LoCoH: Also known as k-NNCH, Fixed k LoCoH is described in
Getz and Willmers (2004). The convex hull for each point is constructed
from the (k-1) nearest neighbors to that point. Hulls are merged together
from smallest to largest based on the area of the hull. This method is
implemented in the function `LoCoH.k`
(2) Fixed r LoCoH: In this case, hulls are created from all points within r
distance of the root point. When merging hulls, the hulls are primarily
sorted by the value of k generated for each hull (the number of points
contained in the hull), and secondly by the area of the hull. This method
is implemented in the function `LoCoH.r`
(3) Adaptive LoCoH: Here, hulls are created out of the maximum nearest
neighbors such that the sum of the distances from the nearest neighbors
is less than or equal to d. Use the same hull sorting as Fixed r LoCoH.
This method is implemented in the function `LoCoH.a`
