---
title: "Day 3 - Preparing covariates for a basic RSF"
author: "Dana Seidel"
date: "January 5, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r, warnings=FALSE, message=FALSE}
#install.packages("lme4")
library(raster)
library(lme4)
library(sf)
library(tidyverse)
library(adehabitatHR)
library(mapview)
```

# Building a RSF
Manual of Applied Spatial Ecology, Chapter 8, Resource Selection. 
http://ecosystems.psu.edu/research/labs/walter-lab/manual/chapter-8-resource-selection

## Step 1: Read in, Clean, and project all your Data
```{r}
zebra <- read_csv("data_files/Zebra_Used_LatLong_Full.csv") %>% 
  dplyr::select(ID = Name, 4:6) %>% 
  st_as_sf(., coords = 3:4, crs = "+init=epsg:4326") %>% 
  st_transform("+init=epsg:32733") %>% 
  mutate(timestamp = lubridate::mdy_hm(Date))

water <- st_read("data_files/ENP shapefiles/functional water.shp") %>% 
  st_transform("+init=epsg:32733")
ggplot(water) + geom_sf()

roads <- st_read("data_files/ENP shapefiles/enp roads.shp", crs = "+init=epsg:4326") %>% 
  st_transform("+init=epsg:32733")
ggplot(roads) + geom_sf()

NDVI <- raster("data_files/NDVI_200903.tif") %>% 
  projectRaster(crs="+proj=utm +zone=33 +south +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
```

### Keep in mind temporal effects!
RSF may be a spatial analysis but if we think selection or availability may 
change over time, we must control for that in our models, either via appropriate 
covariate inclusion, or simpky by limiting our scope. 

```{r}
# NDVI was collected in March 2009... perhaps we should limit to only the rainy season months. 
# Jan, Feb, Mar

zebra %>% 
  filter(lubridate::month(timestamp) %in% 1:3) %>% 
  sf::select.sf(-Date) %>% 
  mutate(Used = 1) -> zebra
```

## Preparing Used v. Available
```{r, screenshot.force=TRUE}
# Choosing an available range. 
# semiconservative option - 99% UD
#zebra_ud <- adehabitatHR::kernelUD(as(zebra, "Spatial"), 
#                                   grid = 450)
zebra_ud <- adehabitatHR::kernelUD(as(zebra, "Spatial")[1],  # add the id column and you get unique homeranges
                                   grid = 450)
zebra_hr <- adehabitatHR::getverticeshr(zebra_ud, 99)

# let's visualize to see
homerange <- mapview(zebra_hr)
homerange

# Note a couple things here: 
# - the pan: is this really available?
# - the park fence?
# we could mask by these polygons."expert opinion"

# For now we are going to use these to sample available points. 
# First I need to know how many points to sample in each home range. 

zebra_ids <- unique(zebra$ID)
availables <- list()
for(i in 1:length(zebra_ids)){
  st_sample(st_as_sf(zebra_hr)[i,], 10*nrow(filter(zebra, ID == zebra_ids[i]))) %>% # sample individual UDs
  #st_sample(st_as_sf(zebra_hr), 10*nrow(filter(zebra, ID == zebra_ids[i])))  %>%  # 1:10 ratio; population UD
  st_sf(geometry = .) %>%
    mutate(ID = zebra_ids[i], 
          timestamp = NA, 
          Used = 0) -> availables[[i]] 
}

availables %>% 
  do.call(rbind,.) %>%
  rbind(zebra, .) -> zebra_all

```

# Preparing linear measures
```{r, screenshot.force=TRUE}
head(roads)
unique(roads$TYPE)

mapview(roads, homerange)
primary_roads <- filter(roads, TYPE %in% c("Tar", "Gravel"))
mapview(primary_roads, homerange)

# next step take 20 sec or so
dist<- st_distance(y=zebra_all, x=primary_roads) # a units matrix dim =[nrow(x), nrow(y)]
road_dist <- apply(dist, 2, min)
head(road_dist)

#
dist<- st_distance(y=zebra_all, x=water) # a units matrix dim =[nrow(x), nrow(y)]
water_dist <- apply(dist, 2, min)
head(water_dist)
```

# Preparing raster covariates
```{r}
zebra_all %>% mutate(NDVI = raster::extract(NDVI, as(., "Spatial")), 
                     dist_rd = road_dist,
                     dist_H2O = water_dist) %>% 
  group_by(ID) -> zebra_full
```

# Building the RSF
## Approaches
From Spatial Ecology 
> There are several ways to to calculate RSFs in R using logistic functions that can assess population level or intra-population variation. The use of General Linear Models with various function using the `lme4` package is often used for estimating population-level models only. Alternatively, we can assess intra-population variation using the `glmer` function. Assessing intra-population variation is a mixed-model approach that provides a powerful and flexible tool for the analysis of balanced and unbalanced grouped data that are often common in wildlife studies that have correlation between observations within the same group or variation among individuals at the same site (Gillies et al. 2006).

### A note on colinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

When preparing for a regression analysis we want to ensure that our covariates are not 
overly correlated. If two variables are highly correlated with one another it makes it hard
to distinguish their individual effects on the response variable - in this case, the use 
of a habitat type. 

Thankfully for us, productivity and distance to road and water do not seem to be especially
correlated in our system:

```{r}
cor(zebra_full$NDVI, zebra_full$dist_rd)
cor(zebra_full$NDVI, zebra_full$dist_H2O)
```

People debate how to deal with multicollinearity but
a standard rule of sum is to to drop variables that have a correlation coeffienct
of over .7. 

### Population Level - glm

```{r}
m <- glm(Used ~ NDVI + dist_rd + dist_H2O, data=zebra_full, family=binomial(link="logit"))

stepAIC(m)
```

##### A Note on Procedure:
Clearly this is a simplified example for trainings sake. Typically in a full RSF
analysis you would evaluate multiple candidate models and use model selection
and averaging procedures to choose the one best supported by your data. Additionally
one would expect to use some cross validation procedure to evaluate the predictive
ability of your best model. 

### Account for individual variation - glmer

```{r}
zebra_full$ID <- as_factor(zebra_full$ID)

mm <- glmer(Used ~ scale(dist_rd) + scale(dist_H2O) + (1|ID), 
           data=zebra_full, family=binomial(link="logit"))

mm
```


# Predicting from an RSF

Often we want to interpolate a selection surface from our best model. To do this
you must ensure first that you have complete data layers across you entire 
area of interest. Since our NDVI layer covers our entire zone we need not recreate 
that, but for our roads and water variables we need to create layers for
distance to roads or distance to water across our entire regions. 
We are going to show you below how to calculate this in the below
chunk but as this takes a long time to process, we will pull in pre-calculated 
geotiff in the next chunk for further mapping. 

```{r, eval=FALSE}

# bring in boundary of area of interest
ENP <- st_read("data_files/ENP shapefiles/enp fence poly.shp") %>% 
  st_union %>% st_transform("+init=epsg:32733")

# Let's match the resolution of our NDVI layer
# res(NDVI)

# make a grid and calculate distances
enp_grid <- st_make_grid(ENP, n = c(1266, 476), what= "centers") 
# convert to raster
dist_raster <- raster(as(enp_grid, "Spatial"), ncol=1266, nrow= 476)

# dist_road_grid <- st_distance(primary_roads, enp_grid)
dist_water_grid <- st_distance(water, enp_grid)
# dist_raster[] = apply(dist_road_grid, 2, min)
dist_raster[] = apply(dist_water_grid, 2, min)


plot(dist_raster)
plot(st_geometry(water), add=TRUE)

# alternative method using sp and regos libraries -- comparable, doesn't require flip
#require(rgeos)
#dd = gDistance(as(water, "Spatial"), as(dist_raster,"SpatialPoints"), byid=TRUE)
#dist_raster[] = apply(dd,1,min)
```

To do the above for both the water and road layers would take 5-10 mins, so instead
let's read in the final rasters for processing:
```{r}
dist_road <- raster("data_files/Dist_PrimaryRoads.tif") %>% flip(2)
dist_water <- raster("data_files/Dist_Water.tif") %>% flip(2)

# why are we using flip?
# these files were created with st_distance function. gDistance would not require this. 
# http://rspatial.org/spatial/rst/8-rastermanip.html
# "flip lets you flip the data (reverse order) in horizontal or vertical direction – typically to correct for a ‘communication problem’ between different R packages or a misinterpreted file."
# sf and raster have a "communication problem", in the way the distance is calculated in st_distance. 



# make a dataframe of all variable points:
# step 1 make ndvi points match in space to the dist variables
ndvi <- resample(crop(NDVI, extent(dist_road)), dist_road)
# stack them all
rasters <- stack(ndvi, dist_road, dist_water)
names(rasters) <- c("NDVI", "dist_rd", "dist_H2O")

#predictions based on best model
best_m <- glm(Used ~ dist_rd + dist_H2O, data=zebra_full, family=binomial(link="logit"))

# make a predicted raster layer based upon glm
predictions = raster::predict(object = rasters, model=m, fun=predict)
predictions = exp(predictions)

plot(predictions)
```




