---
title: "Exploring Movement Data in R"
author: "Dana Seidel & Eric Dougherty"
date: "January 4, 2018"
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(move)
library(adehabitatLT)
```

# Movement data
As mentioned this morning, movement data may come in a variety of scales and resolutions, 
requiring careful consideration before analysis. 

Here we are going to focus on GPS relocation data as it is one of the most common
forms of tracking data and provides a (mostly) consistent record of an animals movement path. 

Justr as there are multiple semi-cooperative packages in R for working with spatial
data generaly, there are many packages in R used to handle movement data, and
an ever growing number built for specific analyses. Unfortunately there is no strict
consensus yet on which class
movement data objects should be in across packages, and researchers often 
have to be comfortable converting from one to the other depending on the task at hand and the 
format required by each package. Although there are most definitely more R packages
created for dealing with animal movement data, we are going to focus today on the two most
prominent `move` and `adehabitatLT`.

To begin we are going to introduce Movebank and `Move` & `MoveStack` objects with the 
library `move`. Later we will introduce the `adehabitatLT` package and demonstrate
how easy it is to pull out primary stepwise characteristics like step length, turning
angle, and net displacement from our trajectories. These primary path characteristics
will form the basis for many more complex movement analyses including our simulations 
later next week. 

## Movebank
Though often researchers in your position
will have data of your own, there has been an effort 
to share movement data from other collections through the Movebank database. Movebank 
is a free online infrastructure created to help researchers manage, share, analyze, 
and archive animal movement data. `move` is an R package that contains functions to
access movement data stored in [movebank.org](http://www.movebank.org) as well
as tools to visualize and statistically analyze animal movement data. 
`move` is addressing movement ecology questions that complement existing efforts 
such as `adeHabitat` and other packages which we will introduce later.

There are 3 ways to load movement data into R as a `move` object:
1. Using the movebank API
2. Directly loading movebank-formatted .csvs
3. Converting data from a traditional dataframe 

Here let's consider some Albatross data, from movebank study #2911040: To load 
data directly from the movebank API like I will here, you would need your own 
movebank username and password, here I
have stored my password in my local environment as the variable `pass`. 

```{r, eval=FALSE}
loginStored <- movebankLogin(username="dpseidel", password=pass)

#search for studies - 
searchMovebankStudies(x="albatross", login=loginStored)
#get study ID - 
getMovebankStudy(2911040, login=loginStored)$citation

#check for animal IDs - 
getMovebankAnimals(2911040, login=loginStored)
#get the all data - getMovebankData()
#get only specific animal by specifying individual_id

albatross <- getMovebank("event", study_id = 2911040, login=loginStored)

write_tsv(albatross, "Study2911040")
# getMovebank("event", login, study_id,...): returns the sensor measurements from a study. 
# See also getMovebankData, getMovebankNonLocationData.
```
To save us the effort of getting you all Movebank logins at this point, I have provided
the tsv. We can use this to demonstrate the 3rd way of loading in movement data.
```{r}
study <- read_tsv("data_files/Study2911040", comment="##",
                  col_types = "Tnncc")
study <- as.data.frame(study)
                       
albatross <- move(x=study$location_long,
              y=study$location_lat,
              time=study$timestamp,
              data=study,
              animal = as.factor(study$individual_id),
              sensor = as.factor(study$tag_id),
              proj=CRS("+proj=longlat"))

head(albatross)
class(albatross)
```
We've created our own MoveStack. We can plot it using the move::plot command and if we wish, `mapview()` which converts it to a `sf` object on the fly. 

```{r}
plot(albatross, type="l") # all of them
```

If we are curious to compare behaviors across individuals, we can split the moveStack, 
into separte move objects with the command `split`
```{r}
ids <- split(albatross)
```

This allows us to plot and manipulate the individuals separately:
```{r}
plot(ids[[1]], type='l')
plot(ids[[28]], type='l')

par(mfrow=c(2,2))
plot(ids[[1]], type='l', main= names(ids[1]))
plot(ids[[2]], type='l', main= names(ids[2]))
plot(ids[[3]], type='l', main= names(ids[3]))
plot(ids[[4]], type='l', main= names(ids[4]))
par(mfrow=c(1,1))
```


And of course, just as before, we can manipulate, and clean up, this data easily when converted to an sf object:
```{r}
albatross_sf <- 
  study %>% na.omit() %>%
  st_as_sf(coords = c("location_long", "location_lat"), crs=4326)
```

For instance, in this state, it's simple to look at how many records we have per individual:
```{r}
albatross_sf %>% 
  group_by(individual_id) %>% 
  tally()

# something we could do on the movestack by calling...
table(albatross@trackId)
```

Or map the paths of a particular pair:
```{r, screenshot.force=TRUE}
albatross_sf %>% 
  filter(individual_id %in% c(2911059,2911062)) %>% 
  mapview(zcol="individual_id")
```

and easily convert it back to a `movestack` as needed. 
```{r}
as(albatross_sf, 'Spatial') %>% as(., "MoveStack") 
```

There is lots more the `move` package has to offer but for now, we are going to move
onto ltraj objects and the primary path characteristics
they make so easy to extract. We will come back to the `move` package tomorrow 
when we discuss methods for estimating home ranges. 

## Ltraj objects 

The class ltraj is intended to store trajectories of animals. 
`ltraj` objects function neatly as lists, similar in some ways 
to the list of Move objects that was created when we split out albatross MoveStack. 
A key difference between Move and ltraj objects is not only in the structure of data
but also the attributes calculated. Ltraj ojects automatically calculate common path
characteristics like step length, relative turning angle, absolute turning angle, and
net squared displacement from which much of subsequent movement modelling builds off of. 
Additionally from these primary path characteristics it's simple to calculate 
secondary statistics like sinuosity,  residence time, or directional persistence. 

`ltrajs` can be created from a traditional data frames, spatial dataframes, or 
directly from move objects using the command `as.ltraj` and, in the case of the 
dataframes, specifying the coordinates. 


```{r}
# from move object
alba_ltraj  <- as(albatross, 'ltraj')

alba_ltraj  
```
Note all the information the this class gives us up front. Not only can we see the 
number of relocations from each of our animals we can see exactly the duration of 
each "burst" and that this is an "irregular" "Type2" trajectory. Just what exactly does that mean?

### Side note: types of trajectory

The adehabitat packages, and ltraj objects, distinguish between 2 types of trajectories:
1. Trajectories of type I are characterized by the fact that the time is not
precisely known or not taken into account for the relocations of the
trajectory (i.e. sampling of tracks in snow)
2. Trajectories of type II are characterized by the fact that the time is
known for each relocation. This type of trajectory is in turn be divided
into two subtypes:
  – regular trajectories: these trajectories are characterized by a constant
time lag between successive relocations;
  – irregular trajectories: these trajectories are characterized by a
variable time lag between successive relocations


It's worth emphasizing that functions in `adehabitatLT` are mainly designed to deal with
type I or type II regular trajectories. Irregular trajectories are harder to analyze as their 
descriptive parameters may not be compared when computed on different time lags.

### Side note: Making an irregular trajectory regular. 

From here on out we are going to work with some ibex data already formatted as ltrajs
and provided by the adehabitatLT package. First we want to load both the *regular* 
trajectory data "ibex" and the raw *irregular* trajectory "ibexraw" just to get a sense of the difference. 

```{r}
data(ibex)
data(ibexraw)

ibex
ibexraw
```

From a simple histogram on the first individual of raw data, we can see there are some big time gaps. 
```{r}
hist(ibexraw[1], "dt", freq = TRUE)
```

Looking at this rawibex data, we can see that the median `dt` is:
```{r}
median(ibexraw[[1]]$dt)
```

From that we can expect that the unit was set to take a fix every 14400 seconds
(4 hrs). To regularize a trajectory like this one that is not too irregular -
we need to round our dt's and insert missing values where needed. Keep in mind, not 
all trajectories can or should be regularized - we will touch on more advanced methods 
for dealling with these next week. 

In this case, we need to make our first timestamp exact and set the missing
values accordingly: 
```{r}
ibex153_raw <- ibexraw[1]
ibex153_raw[[1]]$date <- lubridate::force_tz(ibex153_raw[[1]]$date, "GMT") # hack to simplify timezones. 

## The reference date: the hour should be exact (i.e. minutes=0):
ref <- strptime("00:00:00", "%H:%M:%S", tz="GMT")

ibex153_raw %>% 
  setNA(., ref, 4, units = "hour") %>%  # set the missing values
  sett0(., ref, 4, units = "hour") -> ibex153   # round the fixes to 4 hours. 

is.regular(ibex153)

```

Now our trajectory is regular and we can trust and compare all those primary path characteristics
we care about! 

# Primary Path Characteristics
If we dig into any one of our ltraj list objects, we'll see a dataframe 
containing certain basic characteristics from our paths. Specifically: 

- Step Length (i.e `dist`): the distance between successive relocations is often used in animal
movement analysis (e.g. Root and Kareiva 1984, Marsh and Jones 1988);
```{r}
ibex[[1]]$dist

hist(ibex[[1]]$dist)
```

- Turning Angle (`rel.angle` v. `abs.angle`)

Your relative turning angle measures the change of direction between
the step built by relocations i − 1 and i and the step built by relocations
i and i + 1 (what we generally refer to as “turning angle”). It is often used together with
the parameter dist to fit movement models (e.g. Root and Kareiva 1984,
Marsh and Jones 1988). For comparison the absolute angle is the angle between the x direction and the step
built by relocations i and i + 1. 

```{r}
ibex[[1]]$abs.angle
ibex[[1]]$rel.angle
hist(ibex[[1]]$rel.angle)
```

- Net Squared Displacement, i.e R2n,the squared distance between the first relocation of the trajectory
and the current relocation is often used to test some movements models
(e.g. the correlated random walk, see the seminal paper of Kareiva and
Shigesada, 1983).

```{r}
ibex[[1]]$R2n
hist(ibex[[1]]$R2n)
```

Note these distributions plotted above as they often form the basis for accurate simulations
of animal movemnt paths. 

# Some Secondary Characteristics
From this set of characteristics we can easily pull out secondary path characteristics like: 

- Velocity: 
```{r}
  ibex[[1]]$dist/ibex[[1]]$dt * 3.6 #convert meters per sec to kmph
```

- First Passage Time
```{r}
# Calculate fpt for radii of a kilometer to 5 kilometers
ibex_fpt <- fpt(ibex153, radii = seq(1000, 5000, 1000), units="hours")

# plot variation over time
plot(ibex_fpt, scale=1000)

# plot mean fpt across scales
meanfpt(fpt(ibex153, radii = seq(1000, 5000, 1000), units="hours"))
```


These are just a few of the metrics you might want to derive from your paths. 
Additional secondary characteristics include *straightness*, *sinuosity*, 
*directional persistence*,  *persistence velocity*, *residence time*, 
*time to return*, and more.
We will encounter many of these on later days of this workshop. 

In addition to the derivation of these primary and secondary path cahracteristics, 
adehabitatLT is especially useful for path segementation and some simple movement models. 
We will get more into these analyses at a later date. If interested, I highly recommend
consulting the packages very thourough documentation. 
