---
title: "Projecting & Plotting - Movement Data"
author: "Dana Seidel"
date: "January 4, 2018"
output: github_document
---

- Movebank library (move package)
- adehabitat, ltrajs, path statistics

# Movement data

# Ltraj objects

# Path Metrics 
## Step Length
## Turning Angle (Relative v. Absolute)
## Tortuosity, Straightness Index
## Persistence, Persistence Velocity
## Displacement, Net Squared Displacement




library(movebank)
readRenviron("../../.Renviron")
#study <- movebank("study", id = 2911040)
#individuals <- movebank("individuals", study_id = 2911040)
albatross <- movebank("event", study_id = 2911040)
```

```{r}
albatross_sf <- 
  albatross %>%
  na.omit() %>%
  st_as_sf(coords = c("location_long", "location_lat"), crs=4326)
```

```{r}
albatross_sf %>% group_by(individual_id) %>% tally()
albatross_sf %>% filter(individual_id %in% c(2911059,2911062)) %>% mapview(zcol="individual_id")

```


```{r}
## Leaflet doesn't do "point" and "mutipoint" for `sf` objects directly, but we can combine leaflet and mapview
#albatross_sf %>% filter(individual_id %in% c(2911059,2911062)) %>% leaflet()
data(quakes)
q <- st_as_sf(quakes[1:20,], coords = c("long", "lat"), crs=4326)
# Show first 20 rows from the `quakes` dataset
mapview(q)@map %>% addTiles() %>%
  addMarkers(data = quakes[1:20,], ~long, ~lat, popup = ~as.character(mag), label = ~as.character(mag))
```



----


Consider spatio-temporal data with `stars` package: 

```{r}
#devtools::install_github("r-spatial/stars", update = FALSE)

```

Buffalo movement data from <http://dx.doi.org/10.5441/001/1.j900f88t>


```{r message=FALSE}
#buffalo_meta <- read_csv("https://www.datarepository.movebank.org/bitstream/handle/10255/move.611/Kruger%20African%20Buffalo%2c%20GPS%20tracking%2c%20South%20Africa-reference-data.csv")

buffalo_meta <- read_csv("../data/buffalo-meta.csv")
```

```{r}
#buffalo <- read_csv("https://www.datarepository.movebank.org/bitstream/handle/10255/move.610/Kruger%20African%20Buffalo%2c%20GPS%20tracking%2c%20South%20Africa.csv")

buffalo <- read_csv("../data/buffalo.csv")
buffalo
```

```{r}
buffalo_sf <- st_as_sf(buffalo, coords = c("location-long", "location-lat"), crs=4326)
```

One of the best things about the `sf` library is that we can leverage all of our `dplyr` verbs to manipulate the laters.  For instance, this `buffalo` movement data sure has a lot of records.  Let's subset it to focus only the individual named "Queen" and use the first two months of data:


```{r}
queen <-
buffalo_sf %>% 
  filter(`individual-local-identifier`=="Queen") %>%
  filter(timestamp < as.POSIXct("2005-05-01"))
```

We can now plot this data efficiently:

```{r}
queen %>%
  ggplot() + geom_sf(size=0.2, shape="+")
```

Hmm, it would be helpful to have this data on a map.  Let's take a look at some of the built-in maps:

```{r}
#sa <- map("worldHires","South Africa", plot=FALSE)
st_as_sf(map('world', "South Africa", plot = FALSE, fill = TRUE, xlim = c(14,35), ylim=c(-35,-20))) %>% 
  st_transform(4326) %>%
  ggplot() + geom_sf() 
```

Maybe we want finer resolution:

```{r}
download.file("http://biogeo.ucdavis.edu/data/gadm2.8/shp/ZAF_adm_shp.zip", "sa.shp.zip")
unzip("sa.shp.zip", exdir = "sa.shp")
```



```{r}
south_africa <- st_read("sa.shp")
south_africa %>% 
  ggplot() + geom_sf()
```

```{r}
b <- st_bbox(buffalo_sf)
```

```{r}
kruger <- south_africa %>% filter(NAME_1 %in% c("Limpopo", "Mpumalanga"))

queen %>%
  ggplot()  + 
  geom_sf(data = kruger) +
  geom_sf()  # map first, pts on top
  #coord_sf(xlim = c(31.6,32), ylim = c(-25, -24))
  #coord_sf(xlim = c(b['xmin'], b['xmax']), ylim = c(b['ylim'], b['ymax']))
```

Hmm, that seems like a lot of work.  Can't we just get some standard map layers for our data? And some nice interactive zoom/pan features?

```{r}
mapview(queen, zcol="timestamp")
```

Yeah, that's more like it.

----------


```{r}
p4s <- "+init=epsg:4326 +proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
```

- `sf`: a Simple Feature data.frame
- `sfc`: Simple Feature column, containing the geometry of a simple feature frame.
- `sfg`: a Simple Feature Geometry

Hack cooridnates into WKT format to exploit conversion from CSV

```{r}
buffalo %>%
  rename(X = `location-lat`, Y = `location-long`) %>%
  mutate(WKT = paste0("POINT (", X, " ", Y, ")")) %>%
  select(-X, -Y) %>%
  write_csv("buffalo_st.csv")
buffalo_st <- st_read("buffalo_st.csv")

st_crs(buffalo_st) <- p4s

```


Construct a `sf` data.frame manually? See https://github.com/edzer/sfr/issues/385

```{r}

#buffalo %>% rowwise() %>% summarise(long = `location-long`, lat=`location-lat`)

## better than purrrlyr::by_rows
## need dplyr >= 0.7.0 to support list-columns

buffalo_sf <-
buffalo %>%
  rowwise() %>%
  mutate(geometry = 
           st_geometry(
             st_point(c(`location-long`, `location-lat`)))) %>% 
  ungroup() # remove rowwise class

as(buffalo_sf, "sf")
# 
# class(buffalo_sf) <- c("sf", class(buffalo_sf))

## errors:
#st_crs(buffalo_sf) <- p4s
```

Construct a multipoint geometry. (but, we want 1 geometry per entry!)

```{r}
sfg <- st_multipoint(as.matrix(buffalo[c("location-long", "location-lat")]))
sfc <- st_geometry(sfg)
st_crs(sfc) <- p4s
```

```{r}
library(sp)
buffalo_sp <- SpatialPointsDataFrame(coords = buffalo[c("location-long", "location-lat")], 
                                     data = buffalo, 
                                     proj4string = sp::CRS(p4s))

sf <- as(buffalo_sp, "sf")

#sf2 <- st_as_sf(buffalo_sp) #identical to above
# sfc <- st_as_sfc(buffalo_sp) # identical to st_geometry(sf)
```

Consider Albatross data, study id: 2911040


Look at env var layers from <https://www.movebank.org/node/7471>

http://proj4.org/faq.html
