---
title: "Day 3 - MCP & Utilization Distributions"
author: "Dana Seidel & Eric Dougherty"
date: "January 5, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(messages = FALSE)
```

```{r}
library(tidyverse)
library(sf)
library(adehabitatHR)
library(move)
```

Today we are going to explore 3 main methods for home range estimation in R. We 
will again work primarily with the adehabitatHR library and the move library. Most of this code
and text is an adaptation of the extensive adehabitatHR vignettes, which we encourange you
to consult for further detail and references. 

The adehabitatHR library contains methods to calculate: 

1. The Minimum Convex Polygon (Mohr, 1947);
2. Several kernel home range methods:
  - The “classical” kernel method (Worton, 1989)
  - the Brownian bridge kernel method (Bullard, 1999, Horne et al.
2007);  
  - The Biased random bridge kernel method, also called “movementbased
kernel estimation” (Benhamou and Cornelis, 2010, Benhamou,
2011);
  - the product kernel algorithm (Keating and Cherry, 2009).
3. Several home-range estimation methods relying on the calculation of convex
hulls:
  - The modification by Kenward et al. (2001) of the single-linkage clustering
algorithm;
  - The three LoCoH (Local Convex Hull) methods developed by Getz
et al. (2007)
  - The characteristic hull method of Downs and Horner (2009)
  
Today we will explore the MCP method, the classical and brownian kernal methods, and
Getz's LoCoH methods. Code and comparison across methods are well described 
in the adehabitat vignette and the cited literature. 

# Minimum Convex Polygon 

As dicussed in lecture, the simplest delineation of a home range is an MCP, 
creating the polygon of minimum area around a certain percentage of relocation points. 
The MCP is simple and used widely in ecology. 

If you are curious to estimate the overall area of your animal's home range
the `move` package includes a function to bootstrap the mcp estimation:

```{r}
data(leroy) # a package moveobject
hrBootstrap(x=leroy, rep=25, unin='km', unout='km2')
```
The resulting table and plot can give you a fairly quick estimate of your animals stable home range.
Note that this method works especially well for a territorial animal like a fisher, 
but might appear much less stable for a migratory or nomadic individual.

If, however, you need to delineate the boundaries of the MCP, the adehabitatHR library
has more options for you. The `mcp` function allows you to specify the percentage
of coordinates to be included and works on any two column dataframe specifying the coordinates
of animal relocations:
```{r}
library(adehabitatHR)

data(bear)
xy <- SpatialPoints(na.omit(ld(bear)[,1:2]))

mcp <- mcp(xy, percent=90)

ggplot() + geom_sf(data = st_as_sf(mcp)) + geom_sf(data=st_as_sf(xy))
```

Yikes! look at all that "unused" space contained within even a 90% mcp!

We'll get to better metrics shortly but if you want to compare the area of your mcp across percentages, the `mcp.area` function works well:
```{r}
mcp.area(xy, percent = seq(20,100, by = 5),
         unin = c("m", "km"),
         unout = c("ha", "km2", "m2"), plotit = TRUE)
```
 
If you are curious to see, what's going on under the hood of the adehabitatHR 
mcp functions, I recommend checking out [this blog post](https://www.r-bloggers.com/home-range-estimation-mcp) 
on the subject by Mitchell Gritts. 
         
#  Worton Kernel UD
https://www.r-bloggers.com/adehabitathr-visualization/
https://www.r-bloggers.com/visualizing-utilization-distributions-again/
adehabitatHR
```{r}
kud <- kernelUD(xy, h="href")
image(kud) + title("Bear UD")

## Kernel home range
jj <- kernel.area(kud)                  ## home range size
plot(jj)                                   ## Plots home range size
ver95 <- getverticeshr(kud) ## home-range contours
ver80  <- getverticeshr(kud, percent = 80)
plot(ver95)  + plot(ver80, add=TRUE, col="green")  +  points(xy)   ## Plots contours

```





