---
title: "Slighty More Advanced R"
author: "Dana Seidel & Eric Dougherty"
date: "1/4/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("printr") 
```

In our earlier section, we discussed the basics of importing, manipulating, 
and visualizing vector data in R, specifically focused on using the "tidyverse".
In this section we want to review some additional powerful tools that we will need
as we dive into simulations in R later in the week. Fundamentals in coding such as functions,
loops, creating/representing data in a `data.frame` / `tibble` objects, and random number generation
will be covered here.

# Random Number Generation

For this section, instead of importing a data set, we are going to create our own example data set 
using random number generation. Knowing how to do this can be very powerful when you want to 
test your code, build reproducible examples, or most importantly for our case, build stochastic simulations. 

The base functions necessary for random number generation in R are tied to probability distributions. The 
most commonly used  distributions being:

- Uniform 
- Normal (Guassian)
- Binominal

R is built around, by, and for statistics professionals so there are many other 
distributions (e.g. Poisson, Weibull, Gamma, etc. ) built into the `stats` package. 
We're going to focus on the
most basic distributions for now but should you need a specific obscure one,
know that it's probably already built into R for your use. 

To generate random numbers according to these distributions there are a set of functions,
all with the prefix `r`, i.e. `runif()` `rnorm()`, `rbinom()`

Within these functions you can specify the necessary variables for the specific distribution you want to pull from. 
For example, if we want 10 numbers pulled randomly from between 0 and 2000:

```{r} 
runif(n = 10, min = 0, max = 2000)
```

Now you'll notice that this returns fractional values, if you want whole numbers, the `sample` function can handle this:

```{r}
sample(0:2000, 10, replace=T)

# this is the same as 
sample.int(2000, 10)
```

**Keep in mind, `sample` is an especially powerful function because you can use it to randomly sample
any empirical dataset you have.**

Often when we are simulating a process, we want to pull our numbers not from a uniform distribition 
but instead from a distribution that matches the process we are trying to siumulate. For example, 
if we wanted to simulate coin flips, all we have to do is generate "flips" from the Bernoulli distribution
(otherwise known as a Binomial distribution with size = 1 and p = .5). 

Let's flip this (even) coin 20 times:
```{r}
rbinom(20, 1, .5)
```

Or perhaps we need to pull 10 samples from a process that by the law of large numbers, probably has
a normal distribution with a mean of 50 and a sd of 12, with the `rnorm` function that's simple!

```{r}
rnorm(10, 50, 12)

# want those to be intergers?
round(rnorm(10,50,12))

```

## Build some fake data:

So, hopefully you get a sense of just how useful random numbers can be for simulations. 
Sometime we use random number generation, not for simulation but just to produce
fake datasets to use for testing or reproducible examples. 

As practice, let's try building a data frame of fake temperatures in HongKong 
through the 12 months of the year for 10 years from 2000-2010. 

From a quick web search I can get the [mean temperatures through the year](http://www.holiday-weather.com/hong_kong/averages/) in Celsius:
```
	Jan	Feb	Mar	Apr	May	Jun	Jul	Aug	Sep	Oct	Nov	Dec
	16	18	20	24	26	28	29	29	27	26	22	17
```

Using the normal distribution assuming 2 degrees deviation in all months, let's build a dataframe combining our tidyverse 
functions from earlier and by randomly generating 10 samples around these mean temperatures by month. 

```{r}
temp <- data.frame(YEAR = 2001:2010,
               JAN = rnorm(10, 16, 2), 
               FEB = rnorm(10, 18, 2),
               MAR = rnorm(10, 20, 2),
               APR = rnorm(10, 24, 2),
               MAY = rnorm(10, 26, 2),
               JUN = rnorm(10, 28, 2),
               JUL = rnorm(10, 29, 2),
               AUG = rnorm(10, 29, 2),
               SEP = rnorm(10, 27, 2),
               OCT = rnorm(10, 26, 2),
               NOV = rnorm(10, 22, 2),
               DEC = rnorm(10, 17, 2))

temp
```

Great! Now that we have some data to work with, let's get to building our own functions to play with it. 

# Custom functions
[**Supplemental Reading**](http://r4ds.had.co.nz/functions.html)

Writing your own functions is one way to reduce duplication in your code or to 
create custom models for simulation. 

From the Data Science for R Chapter 19: 

      Functions allow you to automate common tasks in a more powerful and general way
      than copy-and-pasting. Writing a function has three big advantages over using copy-and-paste:

      1. You can give a function an evocative name that makes your code easier to understand.

      2. As requirements change, you only need to update code in one place, instead of many.

      3. You eliminate the chance of making incidental mistakes when you copy and paste 
      (i.e. updating a variable name in one place, but not in another).


As a general rule of thumb, you should consider writing a function if you need to run the same
operations more than twice. I.e. don't go copying and pasting things 3 or more times.  

For instance if I need to convert every concentration in the `temp` dataframe from
Celsius to Farenheight I might consider writing out the conversion equation:

$$F = 1.8*C + 32 $$
into a custom function of my own:
```{r}
celsius_to_faren <- function(C){
1.8*C + 32
}
```

Rather than doing something repetitive like: 
```{r, results='hide'}
1.8*temp$JAN + 32
1.8*temp$FEB + 32
# all the way to...
1.8*temp$DEC + 32
```


Now obviously, this is a very simple example but is useful for showing us the guts
of the R function environment. Notice 3 important things about the structure of our function:

1. all custom functions should be assigned a **name** (i.e. celsius_to_faren). Keep in mind, your
code has 2 audiences, the computer that needs to be able to run it, and the humans
that need to be able to read it (your future self included!). Keep code clean and 
your naming style consistent to facilitate readbility. 

2. functions take flexibly named arguments (i.e. C) in the paratheses before the operation. Our function would run the exact same way if it was written as:
    ```{r}
      celsius_to_faren <- function(x=15){
      1.8*x + 32
      }
    ```
    - Note also that function arguments can be given defaults (e.g. x = 15), allowing 
    them to run with or without that argument newly specified. For example:
    ```{r}
    # it will revert to the default
    celsius_to_faren()
    
    # or we can specify a different value to run the conversion on
    celsius_to_faren(50)
    ```

3. The guts of the function, the operation, goes in between two curly brackets, after the
necessary arguments have been specified. 

As you have seen above, to run our function we simply call it like any other function from R packages:
`celsius_to_faren()`

## Conditions

Often when coding, especially in more complex function, we want to doing an operation
only when a certain condition is met. This is when we need if and if/else statements!

Formal `if` statements have a very similar format to functions:

`
if(condition){
    operation
  }
`

`if` statements have lots of uses both in and outside of custom functions, but for 
now we are going to focus on how to incorporate them into our function writing.

For example we could adapt our above simple function to do something special when 
it encountered a missing value simply by including an `if` statement:

```{r}
celsius_to_faren <- function(x = 15){
  if (is.na(x)){
    return(-999)  # telling the function to return -999 every time it encounters an NA
    }
  1.8*x + 32 # but to still do the regular conversion other times 
  }
```

Now let's throw in a missing value to our table, for testing! Maybe we didn't sample in January of 2002 for some reason:

```{r}
temp[2,2] <- NA

# combine this function with the `mutate` function we learned earlier

temp %>% 
  rowwise %>% 
  mutate(JAN_F = celsius_to_faren(JAN)) %>% 
  select(YEAR, JAN, JAN_F)

# note! the `rowwise` command works to group the dataframe by each row. 
# because our function is written to take a single value, not a full vector, 
# we need to pass our dataframe through `rowwise` first. 

# 2nd note! breaking up your code at each pipe helps with readability of each step.
```

One can string a bunch of `if` statements together using the nested `if/else` structure:

```{r, eval=FALSE}
if (test_expression1) {
   statement1
} else if (test_expression2) {
   statement2
} else if (test_expression3) {
   statement3
} else
   statement4
```

Alternatively, R has a nifty `ifelse()` function that simplifies this into one line. 

`ifelse(test, yes, no)`

For instance, adapting our above function directly into a mutate command:

```{r}
temp %>% 
  mutate(JAN_F = ifelse(is.na(JAN), -999, JAN*1.8 + 32)) %>% 
  select(YEAR, JAN, JAN_F) 
```

# Loops
*Much of this section is modelled after the iteration chapter of R for Data Science. 
Read more in [**Supplemental Reading**](http://r4ds.had.co.nz/iteration.html)*

Loops are used in programming to repeat a specific block of code. Though loops are
used less often in R than many other programming languages, the loops we
cover today are very important for learning to simulate movement 
paths later in the workshop. 

## For loops
In R, For loops are used to repeat a chunk of code, over all the values held inside the given vector. 

There are 3 parts to a simple for loop:

1. The **initialization**: before you start the loop, you need a place to put the results. 
2. The **iteration** vector: this part, i.e. `(i in 1:10)` determines what values we loop over. In this case, the loop would run over 10 different values, 1,2,3,4,5,6,7,8,9,10 
3. The **body** of the loop: the meat of it - the code that actually does the work. Whatever we want to calculate over all those different numbers, goes here. 


So for instance, maybe we want to calculate the annual mean temperature for each of our years of data. 
We can do this with a for loop! 

```{r}
# first we initialize!
averages  <- vector()      # an empty vector for our averages to go

# the we iterate:
for(i in 1:10){                 # take the mean of row i of temp dataframe and put it the averages vector
  averages[i] <- mean(as.numeric(temp[i,2:13]), na.rm=TRUE)       #drop Nas, and drop column 1 to leave out year.
}                                             

# Now we can read out results
averages
```

Once you understand the basic for loop it's useful to consider some important variations, namely:

1. Modifying an existing object, instead of creating a new object.
2. Looping over names or values, instead of indices.
3. Handling sequences of unknown length ("while" loops, covered in the next section)

Consider the following loops and their outputs for examples of these variations, 
for more information see the readings. Note the consitent syntax of these loops. 

```{r}
# Modifying an existing object

jan <- as.vector(temp$JAN)  # make a simple vector of January column

for(i in seq(1,5,1)){
  jan <- append(jan, rnorm(1,16,2))     # add 5 more years temperatures, 1 at a time. 
}

jan


# Or maybe we want to round every column to integers

for(i in 1:length(temp)){
  temp[i] <- round(temp[i])
}

temp
```

```{r}
# Looping over names or indices rather than sequences.

# print the 5th row of just a few specific columns
months <- c("JAN", "MAY", "SEP")
for (i in months){
 print(mean(temp[,i]), na.rm=TRUE)
}
```

## While loops

In a combination of the conditionals introduced earlier and the for loops we saw above, 
in R programming, while loops are used to loop until a specific condition is met.
These can be especially useful if you don't know the length of the input sequence needed 
for you task: a common occurence when doing simulations. 

For example, you might want to loop until you get three heads in a row. 
You can’t do that sort of iteration with the for loop. Instead, you can use a while loop. 
A while loop is simpler than for loop because it only has two components, a condition and a body:

The syntax of a while expression is as follows: 
```{r, eval=FALSE}
while (condition)
{
   body
}
```

Here, `condition` is evaluated first. If the condition is TRUE, than the body of 
the loop is entered and the calculations inside the loop are executed. Once executed, 
the flow returns to evaluate the `condition` for the next element. This is repeated 
each time until the `condition` evaluates to FALSE, in which case, the loop exits.


So, for example, a while loop using a random number generator to simulate flips of an even coin: 

How many flips does it take to get 3 heads in a row?
```{r}
flip <- function(){ # randomly sample 1 "flip" from a bernoulli dist. 
  rbinom(1, 1, .5)
  } 

flips <- 0        # intialize
nheads <- 0

while (nheads < 3) {  # continue loop until nheads is 3. 
  if (flip() == 1) {  # conditional if else statement.
    nheads <- nheads + 1  # IF the flip is a 1 (head) add it to the nheads vector
  } else {            #  If NOT, reset the nheads vector to 0
    nheads <- 0
  }
  flips <- flips + 1  # track the number of flips by adding each iteration to flip vector
}

flips
```

## The `apply` and `map` functions

This same pattern of looping over a vector, in other words, implementing some 
computation or function over 
many elements is so common both baseR and tidyverse have families of functions to 
do it for you: the apply family of functions (baseR) 
or the map family of functions from the purrr library in (baseR)

These functions make iteration much simpler and your code a lot more concise and
readable but may take some time to wrap your head around. We introduce them here just for you 
to be aware of but if you're just starting out, you may find that for loops 
are easier for you to understand. That's okay! Loops work just as well, they 
just take more time and text to code.  

### Apply: Base R

`lapply()` takes arguments `x`, and `FUN` as inputs. `x` is equivalent to the vector you would 
iterate over in a for loop and `FUN` is equivalent to whatever you would put into
the body of the for loop. `lapply` always returns a list. 

So for instance, let's try applying a custom function using `lapply` to take our temp back to celsius. Note that the output is a list. 

```{r, warnings = FALSE}

 celsius_to_faren <- function(x=15){
      1.8*x + 32
 }

lapply(temp[,2:13], celsius_to_faren)

# we can use bind_rows(), to bind this list to a dataframe
lapply(temp[,2:13], celsius_to_faren) %>% bind_rows()
```


You may also see, `sapply()` or `vapply()`, these do the same thing but return the output
in different formats. 

In a slightly different flavor, `apply()` takes the additional `MARGIN` argument, 
which allows you to specify, when x is a matrix
or dataframe, wether you want a function to be run over just the rows, `1`, just the columns, `2`,
or both, `c(1,2)`. 

For example, maybe we want the mean temperature of each year:

```{r}
apply(temp[2:13], 1, mean, na.rm=TRUE)  #drop Nas, and drop column 1 to leave out year. 
```

Or instead the average by month:
```{r}
apply(temp[2:13], 2, mean, na.rm=TRUE)  #drop Nas, and drop column 1 to leave out year. 
```


### Map: tidyverse and the purrr package

The `purrr` package provides a similar family of functions to do this same iterative computation
with the added clarity of providing one funtion each type of output and the benefit of playing
nicely with all the other tidyverse functions
we introduced earlier e.g. `group_by` & the pipe operator `%>%`. 

Just as before, each function takes a vector as input, applies a function to each piece, 
and then returns a new vector that’s the same length (and has the same names) as the input. 
The type of the vector is determined by the suffix to the map function.

- map() makes a list.  (equivalent to `lapply`)
- map_df() makes a dataframe. 
- map_lgl() makes a logical vector.
- map_int() makes an integer vector.
- map_dbl() makes a double vector.
- map_chr() makes a character vector.

Take for example the same operation using between `map`, `lapply`, & `map_df`
```{r}
lapply(temp[,2:13], celsius_to_faren)

map(temp[,2:13], celsius_to_faren)

map_df(temp[,2:13], celsius_to_faren)
```

To do the same thing with for loops requires the following code (& nesting!):
```{r}
faren <- data_frame()
for (i in 2:length(temp)){
  for (j in 1:nrow(temp)){
    faren[j,i-1] <- celsius_to_faren(temp[j,i])
  }
}

names(faren) <- names(temp)[2:13]
faren
```

Makes `map_df` seem pretty handy right?!


